Static Analysis by Path Focusing

### 2011-04-22 report/report.tex
update in the report

\part{Abstract Interpretation: state of the art}

\section{Abstract Interpretation}

@@

\part{Path Focusing technique}

@@

	We construct the $\rho$ formula as follow:

@@

	$$\rho \wedge b_i^s \wedge \bigwedge_{j\neq i,\ j\in P_R} (\neg b_j^s)
	\wedge x_i \in X_{p_i} \wedge \bigvee_{j/(p_i,p_j)\in E} (b_2^d \wedge \neg
	(x_j \in X_{p_j}))$$

	\begin{itemize}
	\item we search a path starting in $p_i$: $b_i^s$ has to be \emph{true}, and
	all other source points are \emph{false}.
	\item the vector $x_i$ of the numerical variables at point $p_i$ is in our
	abstract value $X_{p_i}$
	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
	vector of the numerical variables $x_j$ is not yet included in the current
	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
	conjunction of the inequalities given by $X_{p_j}$ is \emph{false}.
	\end{itemize}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly do some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{algorithm}
\caption{Transitive closure with implicit transition system}\label{gulwani}
\begin{algorithmic}[1] 
\Procedure{TransitiveClosureImplicit}{$p$}
\ForAll {$j \in \{1,..,m\} \setminus \{\delta\}$}
	\State $X_{j} \gets \perp$
\EndFor
\State $X_{\delta} \gets Id$
\While {true}

@@

	\If {$res = unsat$}
		\State \textbf{break}
	\EndIf
	\State Compute $\tau_{p,i}$ from $res$ 
	\State $j$ is the number such as $B_j=true$
	\State $X_{p,\sigma(j,i)} \gets X_{p,\sigma(j,i)} \sqcup
	\tau_{p,i}(X_{p,j})$
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$.

\part{Implementation}

@@

% We need layers to draw the block diagram
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

@@

\begin{figure}[!h]
\centering
\begin{tikzpicture}
    \node (IR) [IR] {Internal Representation\\ (IR)};
    % Note the use of \path instead of \node at ... below. 
    \path (IR.130)+(-\leftdist,0) node (c) [sensor] {C};
    %\path (IR.-150)+(-\leftdist,0) node (cpp) [sensor] {C++};
	\node [sensor] (cpp) [below of=c, yshift=-0.3cm] {C++};
	\node  (lang) [below of=cpp] {$\vdots$};

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in different languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer works directly on the
internal representation (IR), which is a graph representation of
the program. The control flow graph of the program can easily be extracted from
this representation.

\begin{figure}[!h]
\begin{minipage}[c]{.46\linewidth}
\begin{C}
int f(int n) {
	int x;
	int y = n;
	if (n > 0) {
		x = 50;
		y++;
	} else {
		x = 10;
		y--;
	}
	return x+y;
}
\end{C}
\end{minipage}
\begin{minipage}[c]{.46\linewidth}
\begin{LLVM}
define i32 @f(i32 %n) nounwind {
  %1 = icmp sgt i32 %n, 0
  br i1 %1, label %2, label %4

; <label>:2                                       ; preds = %0
  %3 = add nsw i32 %n, 1
  br label %6

; <label>:4                                       ; preds = %0
  %5 = add nsw i32 %n, -1
  br label %6

; <label>:6                                       ; preds = %4, %2
  %x.0 = phi i32 [ 50, %2 ], [ 10, %4 ]
  %y.0 = phi i32 [ %3, %2 ], [ %5, %4 ]
  %7 = add nsw i32 %x.0, %y.0
  ret i32 %7
}
\end{LLVM}
\end{minipage}
\caption{Example of C program and its internal representation in SSA form}
\end{figure}
\FloatBarrier

@@

		The analyzer is implemented with the APRON library \cite{JM09}.
		% live variables
		To compute the set of live variables in SSA: \cite{Boi10}

>>>>

Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
satefy critical systems, such as avionics.

Abstract Interpretation is a general framework used for static analysis.
Linear Relation Analysis (LRA) is a direct application of this framework, that
computes an upper-approximation of the set of the possible states of a numerical
program. The state of a program is defined by the position where we are in the
code, and the actual values of the different numerical variables.
The set of possible states is expressed as a least fixpoint of a set of
equations defining the semantic of the program.

A fundamental fact in static analysis is that it cannot be perfectly precise:
either the set of possible states we compute is unsound (it doesn't contain all
the possible states during execution), either it is incomplete (it contains
states that can never be reached during execution). 
Abstract Interpretation is incomplete, since it always computes an
upper-approximation of the set of possible states.

Linear Relation Analysis approximates the set of possible states as a convex
polyhedron where the dimensions are the numerical variables of the program.
Instead of convex polyhedra, one could choose intervals, octagons, etc.
The fixpoint equation is computed iteratively; that is, successive
approximations of the set of reachable states are computed until they converge
to a fixpoint.

Satisfiability Modulo Theory (SMT) solving is a field in computer science that
aims to decide the satisfiability of a logic formula containing boolean
predicates and elements of a theory, such as linear arithmetic relations between
integers.
In this report, we present a novel refinment of the Abstract Interpretation
technique, that uses SMT-solving to guide the fixpoint iterations such as it
temporarily focus on a selected path in the control flow graph (the graph
expressing the semantic of the program), in order to
obtain more precise results at reasonable cost.

\subsection*{Contribution and Organization}

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such as we can compare precision and cost of the these two methods.
\end{itemize}

The report is organized as follows:
Part \ref{stateoftheartpart} introduces the state of the art in Abstract
Interpretation and more specifically in Linear Relation Analysis.
Part \ref{pathfocusingpart} describes the two techniques
\emph{Lookahead Widening} and \emph{Path Focusing}.
Part \ref{implementationpart} presents the implementation of these two
techniques into an analyzer, and the experimental results we obtained.

\part{Abstract Interpretation: state of the art}
\label{stateoftheartpart}
\section{Abstract Interpretation}

@@

\part{Path Focusing technique} \label{pathfocusingpart}

@@

	We construct the $\rho$ formula as follows:

@@

	$$\rho \wedge b_i^s \wedge \bigwedge_{j\neq i,\ j\in P_R} (\neg b_j^s)
	\wedge x_i \in X_{p_i} \wedge \bigvee_{j/(p_i,p_j)\in E} (b_2^d \wedge \neg
	(x_j' \in X_{p_j}))$$

	where $x_i$ is the vector of the SSA-variables at the beginning of the path,
	and $x_j'$ is the vector of the SSA-variable at the end: in this second
	vector, $\Phi$-variables defined in $p_j$ are primed.

	\begin{itemize}
	\item we search a path starting in $p_i$: $b_i^s$ has to be \emph{true}, and
	all other source points are \emph{false}.
	\item the vector $x_i$ of the numerical variables at point $p_i$ is in our
	abstract value $X_{p_i}$
	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
	vector of the numerical variables $x_j'$ is not yet included in the current
	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
	conjunction of the inequalities given by $X_{p_j}$ is \emph{false}.
	\end{itemize}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	To Be Completed\dots

@@

\begin{algorithm}[!h]
\caption{Transitive closure with implicit transition system}\label{gulwani2}
\begin{algorithmic}[1] 
\Procedure{TransitiveClosureImplicit}{$p$}
\ForAll {$j \in \{1,..,m\} \setminus \{\delta\}$}
	\State $X_{j} \gets \perp$
\EndFor
\State $X_{\delta} \gets Id$
\While {true}

@@

	\If {$res = unsat$}
		\State \textbf{break}
	\EndIf
	\State Compute $\tau_{p,i}$ from $res$ 
	\State Take $j \in \{ k | B_k = true\}$ 
	\State $X_{p,\sigma(j,i)} \gets X_{p,\sigma(j,i)} \sqcup
	\tau_{p,i}(X_{p,j})$
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
compute this number of path at the beginning of the algorithm.

\part{Implementation}\label{implementationpart}

@@

\subsection{LLVM internal representation}
% We need layers to draw the block diagram
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

@@

\begin{figure}[!h]
\centering
\begin{tikzpicture}
    \node (IR) [IR] {\small Internal\\Representation\\ (IR)};
    % Note the use of \path instead of \node at ... below. 
    \path (IR.130)+(-\leftdist,0) node (c) [sensor] {C};
    %\path (IR.-150)+(-\leftdist,0) node (cpp) [sensor] {C++};
	\node [sensor] (cpp) [below of=c, yshift=-0.3cm] {C++};
	\node  (lang) [below of=cpp] {$\vdots$};

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in different languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer works directly on the
internal representation (IR), which is a graph representation of
the program. The control flow graph can easily be extracted from
this representation.

\begin{figure}[!h]
\begin{minipage}[c]{.45\linewidth}
\begin{C}
int main() {
   int x = 0;
   int y = 0;

   while (1) {
      if (x <= 50) y++;
      else y--;

      if (y < 0) break;
      x++;
   }
}
\end{C}
\end{minipage}
\begin{minipage}[c]{.45\linewidth}
\includegraphics[width=8.5cm]{"images/cfg_gopan"}
\end{minipage}
\caption{Our running example and its internal representation in SSA form}
\end{figure}
\FloatBarrier

\subsection{Drawbacks}

During the implementation, we encountered problems due to the use of the LLVM
internal representation. Indeed, this format is not provided for static
analysis, but for code generation. Then, some informations we need to do precise
static analysis are lacking.

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is inexistant in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, one need to
have this distinction to improve precision:
for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
	abstract value at the beginning of the analysis.
	This lack of precision may be recovered by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

@@

		The analyzer is implemented with the APRON library \cite{JM09}, which
		is a common interface to libraries implementing all the features
		for several abstract domains, such as convex polyhedra (NewPolka), 
		octagons (OCT), intervals (BOX).
		In our case, we used the convex polyhedra library, but we could switch
		to octagons or intervals easily.

		\subsection{Dimensions of the Abstract values}
		To compute the set of live variables in SSA: \cite{Boi10}
